--- 
title: "Simulating species abundance data with coenocliner"
status: publish
layout: post
published: true
type: post
tags:
- coenocliner
active: blog
category: R
---

```{r, echo = FALSE, include = FALSE, cache = FALSE}
render_jekyll()
opts_knit$set(progress = TRUE, base.url = "{{ site.url }}/assets/img/posts/", base.dir = "/home/gavin/work/web/jekyll/blog/assets/img/posts/")
opts_chunk$set(results = "hold",
               fig.path = "simulating-species-abundance-data-with-the-coenocliner-package-", 
               comment = NA,
               prompt = FALSE,
               cache = TRUE,
               cache.path = "../_knitr-cache/simulating-species-abundance-data-with-the-coenocliner-package-", 
               fig.show = "hold",
               dev = "png",
               dpi = 85,
               tidy = FALSE,
               fig.height = 7,
               fig.width = 7,
               dev.args = list(pointsize = 10))
opts_chunk$set(fig.lp = "fig:")
```

Coenoclines are, according to the Oxford Dictionary of Ecology 
[@Allaby1998], *"gradients of communities (e.g. in a transect from the 
summit to the base of a hill), reflecting the changing importance, 
frequency, or other appropriate measure of different species 
populations"*. In much ecological research, and that of related fields, 
data on these coenoclines are collected and analyzed in a variety of 
ways. When developing new statistical methods or when trying to 
understand the behaviour of existing methods, we often resort to 
simulating data with known pattern or structure and then torture 
whatever method is of interest with the simulated data to tease out how 
well methods work or where they breakdown. There's a long history of 
using computers to simulate species abundance data along coenoclines 
but until recently no **R** packages were available that performed 
coenocline simulation. **coenocliner** was designed to fill this gap, 
and today, the package was [released to CRAN](http://cran.r-project.org/web/packages/coenocliner/index.html).

**coenocliner** can simulate species abundance or occurrence data along 
one or two gradients from either a Gaussian or generalised beta 
response model. Parameters for the response model are supplied for each 
species and parameterised species repsonse curves along the gradients 
are returned. Simulated abundance or occurrence data can be produced by 
sampling from one of several error distributions which use the 
parameterised species response curves as the expected count or 
probability of occurrence for the chosen error distribution. The 
available error distributions are

 * Poisson
 * Negative binomial
 * Bernoulli (occurrence; Binomial with denominator \\(m = 1\\))
 * Binomial (counts with specified denominator \\(m\\))
 * Beta-binomial
 * Zero-inflated Poisson (ZIP)
 * Zero-inflated negative binomial (ZINB)

You can find the [source code on 
github](https://github.com/gavinsimpson/coenocliner/) and 
[report](https://github.com/gavinsimpson/coenocliner/issues) any bugs 
or issues there. In the remainder of this posting I give an overview of **coenocliner** and show three examples illustrating features of package.

## Introduction to coenocliner

To begin, load **coenocliner** and check the start-up message to see 
if you are using the current (0.1-0) release of the package

```{r, load-library}
library("coenocliner")
```

The main function in **coenocliner** is `coenocline()`, which provides 
a relatively simple interface to coenocline simulation allowing 
flexible specification of gradient locations and response model 
parameters for species. Gradient locations are specified via argument 
`x`, which can be a single vector, or, in the case of two gradients, a 
matrix or a list containing vectors of gradient values. The matrix 
version assumes the first gradient's values are in the first column and 
those for the second gradient in the second column

```{r, show-matrix-grad-locs}
xy <- cbind(x = seq(from = 4, to = 7, length.out = 100),
            y = seq(from = 1, to = 100, length.out = 100))
```

Similarly, for the list version, the first component contains the 
values for the first gradient and the second component the values for 
the second gradient

```{r, show-list-grad-locs}
xy <- list(x = seq(from = 4, to = 6, length.out = 100),
           y = seq(from = 1, to = 100, length.out = 100))
```

The species response model used is indicated via the `responseModel` 
argument; available options are `"gaussian"` and `"beta"` for the 
classic Gaussian response model and the generalise beta response model 
respectively. Parameters are supplied to `coenocline()` via the 
`params` argument. `showParams()` can be used list the parameters for 
the desired response model. The parameters for the Gaussian response 
model are

```{r, show-gaussian-params}
showParams("gaussian")
```

As indicated, some parameters are only supplied once per species, 
regardless of whether there are one or two gradients. Hence for the 
Gaussian model, the parameter `h` is only supplied for the first 
gradient even if two gradients are required.

Parameters are supplied as a matrix with named columns, or as a list 
with named components. For example, for a Gaussian response for each of 
3 species we could use either of the two forms

```{r, param-example}
opt <- c(4,5,6)
tol <- rep(0.25, 3)
h <- c(10,20,30)
parm <- cbind(opt = opt, tol = tol, h = h)     # matrix form
parl <- list(opt = opt, tol = tol, h = h)      # list form
```

In the case of two gradients, a list with two components, one per 
gradient, is required. The first component contains parameters for the 
first gradient, the second element contains those for the second 
gradient. These components can be either a matrix or a list, as 
described previously. For example a list with parameters supplied as 
matrices

```{r, param-example-2}
opty <- c(25, 50, 75)
tol <- c(5, 10, 20)
pars <- list(px = parm,
             py = cbind(opt = opty, tol = tol))
```

Note that parameter \\(h\\) is not specified in the second set as this 
parameter, the height of the response curve at the gradient optima, 
applies globally --- in the case of two gradients, \\(h\\) refers to 
the height of the bell-shaped curve at the bivariate optimum.

Notice also how parameters are specified at the species level. To 
evaluate the response curve at the supplied gradient locations each set 
of parameters needs to be repeated for each gradient location. 
Thankfully `coenocline()` takes care of this detail for us.

Additional parameters that may be needed for the response model but 
which are not specified at the species level are supplied as a list 
with named components to argument `extraParams`. An example is the 
correlation between Gaussian response curves in case of two gradients. 
This, unfortunately, means that a single correlation between response 
curves applies to all species[^1], and is caused by a poor choice of 
implementation. Thankfully this is relatively easy to fix, which will 
be done for version 0.2-0 along with a fix for a similar issue relating 
to the statement of additional parameters for the error distribution 
used (see below).

[^1]: This is not strictly true as you can work out how the species 
parameters are replicated relative to gradient values and hence pass a 
vector of the correct length with the species-specific values included. 
Study the outputs from `expand()` when supplied gradient locations and 
parameters to work out how to specify `extraParams` appropriately

To simulate realistic count data we need to sample *with error* from 
the parameterised species response curves. Which of the distributions 
(listed earlier) is used is specified via argument `countModel`; 
available options are

```{r, count-models, echo = FALSE}
c("poisson", "negbin", "bernoulli", "binary", "binomial", 
  "betabinomial", "ZIP", "ZINB")
```

Some of these distributions (all bar `"poisson"` and `"bernoulli"`) 
require additional arguments, such as the \\(\alpha\\) parameter for 
(one parameterisation of) the negative binomial distribution. These 
arguments are supplied as a list with named components. Again, due to 
the same implementation snafu as for `extraParams`, such parameters act 
globally for all species[^2].

[^2]: Again, this is not strictly true as you can work out how the 
species parameters are replicated relative to gradient values and hence 
pass a vector of the correct length with the species-specific values 
included. Study the outputs from `expand()` when supplied gradient 
locations and parameters to work out how to specify `countParams` 
appropriately

The final argument is *expectation*, which defaults to `FALSE`. When 
set to `TRUE`, simulating species counts or occurrences with error is 
skipped and the values of the parameterised response curve evaluated at 
the gradient locations are returned. This option is handy if you want 
to look at or plot the species response curves used in a simulation.

## Example usage

In the next few sections the basic usage of `coenocline()` is 
illustrated.

### Gaussian responses along a single gradient

This example, of multiple species responses along a single 
environmental gradient, illustrates the simplest usage of 
`coenocline()`. The example uses a hypothetical pH gradient with 
species optima drawn at random uniformally along the gradient. Species 
tolerances are the same for all species. The maximum abundance of each 
species, \\(h\\), is drawn from a lognormal distribution with a mean of 
~20 (\\(e^3\\)). This simulation will be for a community of 20 species, 
evaluated at 100 equally spaced locations. First we set up the 
parameters

```{r, example1-params}
set.seed(2)
M <- 20                                    # number of species
ming <- 3.5                                # gradient minimum...
maxg <- 7                                  # ...and maximum
locs <- seq(ming, maxg, length = 100)      # gradient locations
opt  <- runif(M, min = ming, max = maxg)   # species optima
tol  <- rep(0.25, M)                       # species tolerances
h    <- ceiling(rlnorm(M, meanlog = 3))    # max abundances
pars <- cbind(opt = opt, tol = tol, h = h) # put in a matrix
```

As a check, before simulating any count data, we can look at the 
coenocline implied by these parameters by returning the expectations 
only from `coenocline()`

```{r, example1-expectations}
mu <- coenocline(locs, responseModel = "gaussian", params = pars,
                 expectation = TRUE)
```

This returns a matrix of values obtained by evaluating each species 
response curve at the supplied gradient locations. There is one column 
per species and one row per gradient location

```{r, example1-head-expectations}
class(mu)
dim(mu)
head(mu[, 1:6])
```

A quick way to visualise the parameterised species response is to use 
`matplot()`[^3]

[^3]: until such a time as the **coenocliner** has a `plot` method...

```{r, example1-plot-expectations, fig.cap = "Figure 1: Gaussian species response curves along a hypothetical pH gradient", fig.height = 5}
matplot(locs, mu, lty = "solid", type = "l", xlab = "pH", ylab = "Abundance")
```

The resultant plot is shown in Figure 1.

As this looks OK, we can simulate some count data. The simplest model 
for doing so is to make random draws from a Poisson distribution with 
the mean, \\(\lambda\\), for each species set to value of the response 
curve evaluated at each gradient location. Hence the values in `mu` 
that we just created can be thought of as the expected count per 
species at each of the gradient locations we are interested in. To 
simulate Poisson count data, use `expectation = FALSE` or remove this 
argument from the call. To be more explicit, we should also state 
`countModel = "poisson"`[^4].

[^4]: `countModel = "poisson"` is the default so this can be excluded 
from the call.

```{r, example1-sim}
simp <- coenocline(locs, responseModel = "gaussian", params = pars,
                   countModel = "poisson")
```

Again, `matplot` is useful in visualizing the simulated data

```{r, example1-plot-simulations, fig.cap = "Figure 2: Simulated species abundances with Poisson errors from Gaussian response curves along a hypothetical pH gradient", fig.height = 5}
matplot(locs, simp, lty = "solid", type = "p", pch = 1:10, cex = 0.8,
        xlab = "pH", ylab = "Abundance")
```

The resultant plot is shown in Figure 2 above.

Whilst the simulated counts look reasonable and follow the response 
curves in Figure \ref{fig:example1-plot-expectations} there is a 
problem; the variation around the expected curves is too small. This is 
due to the error variance implied by the Poisson distribution 
encapsulating only that variance which would arise due to repeated 
sampling at the gradient locations. Most species abundance data exhibit 
much larger degrees of variation than that shown in 
Figure \ref{fig:example1-plot-simulations}. 

A solution to this is to sample from a distribution that incorporates 
additional variance or *overdispersion*. A natural partner to the 
Poisson that includes overdispersion is the negative binomial. To 
simulate count data using the negative binomial distribution we must 
alter `countModel` and supply the overdispersion parameter $\alpha$ to 
use[^5] via `countParams`.

[^5]: Recall that this is only easily specifiable globally in version 
0.1-0 of **coenocliner**.

```{r, example1-nb-sim}
simnb <- coenocline(locs, responseModel = "gaussian", params = pars,
                    countModel = "negbin", countParams = list(alpha = 0.5))
```

Using `matplot` it is apparent that the simluated species data are now 
far more relalistic (Figure 3)

```{r, example1-plot-nb-simulations, fig.cap = "Figure 3: Simulated species abundance with negative binomial errors from Gaussian response curves along a hypothetical pH gradient", fig.height = 5}
matplot(locs, simnb, lty = "solid", type = "p", pch = 1:10, cex = 0.8,
        xlab = "pH", ylab = "Abundance")
```

### Generalised beta responses along a single gradient

In this example, I recreate figure 2 in @Minchin1987 and then 
simulate species abundances from the species response curves. The 
species parameters for the generalised beta response for the six 
species in @Minchin1987 are

```{r, example2-beta-pars}
A0    <- c(5,4,7,5,9,8) * 10               # max abundance
m     <- c(25,85,10,60,45,60)              # location on gradient of modal abundance
r     <- c(3,3,4,4,6,5) * 10               # species range of occurence on gradient
alpha <- c(0.1,1,2,4,1.5,1)                # shape parameter
gamma <- c(0.1,1,2,4,0.5,4)                # shape parameter
locs  <- 1:100                             # gradient locations
pars  <- list(m = m, r = r, alpha = alpha,
              gamma = gamma, A0 = A0)      # species parameters, in list form
```

To recreate figure 2 in @Minchin1987 evaluations at the chosen 
gradient locations, `locs`, of the parameterised generalised beta are 
required and can be generated by passing `coenocline()` the gradient 
locations and the chosen species parameters as before, choosing the 
generalised beta response model and using `expectation = TRUE`

```{r, example2-beta-expectations}
mu <- coenocline(locs, responseModel = "beta", params = pars, expectation = TRUE)
```

As before `mu` is a matrix with one column per species

```{r, example2-head}
head(mu)
```

and as such we can use `matplot()` to draw the species responses

```{r, example2-beta-plot-expectations, fig.cap = "Figure 4: Generalised beta function species response curves along a hypothetical environmental gradient recreating Figure 2 in Minchin (1987).", fig.height = 5}
matplot(locs, mu, lty = "solid", type = "l", xlab = "Gradient", ylab = "Abundance")
```

Figure 4 is a good facsimile of figure 2 in @Minchin1987.

### Gaussian response along two gradients

In this example I illustrate how to simulate species abundance in an environment comprising two gradients. Parameters for the simulation are defined first, including the number of species and samples required, followed by definitions of the gradient units and lengths, species optima, and tolerances for each gradient, and the maximal abundance (\\(h\\)).

```{r, example3-params}
set.seed(10)
N <- 30                                           # number of samples
M <- 20                                           # number of species
## First gradient
ming1 <- 3.5                                      # 1st gradient minimum...
maxg1 <- 7                                        # ...and maximum
loc1 <- seq(ming1, maxg1, length = N)             # 1st gradient locations
opt1 <- runif(M, min = ming1, max = maxg1)        # species optima
tol1 <- rep(0.5, M)                               # species tolerances
h    <- ceiling(rlnorm(M, meanlog = 3))           # max abundances
par1 <- cbind(opt = opt1, tol = tol1, h = h)      # put in a matrix
## Second gradient
ming2 <- 1                                        # 2nd gradient minimum...
maxg2 <- 100                                      # ...and maximum
loc2 <- seq(ming2, maxg2, length = N)             # 2nd gradient locations
opt2 <- runif(M, min = ming2, max = maxg2)        # species optima
tol2 <- ceiling(runif(M, min = 5, max = 50))      # species tolerances
par2 <- cbind(opt = opt2, tol = tol2)             # put in a matrix
## Last steps...
pars <- list(px = par1, py = par2)                # put parameters into a list
locs <- expand.grid(x = loc1, y = loc2)           # put gradient locations together
```

Notice how the parameter sets for each gradient are individual matrices 
which are combined in a list, `pars`, ready for use. Also different 
this time is the `expand.grid()` call which is used to generate all 
pairwise combinations of the locations on the two gradients. This has 
the effect of creating a coordinate pair on the two gradients at which 
we'll evaluate the response curves. In effect this creates a grid of 
points over the gradient space.

Having set up the parameters, the call to `coenocline()` is the same as 
before, except now we specify a degree of correlation between the two 
gradients via `extraParams = list(corr = 0.5)`

```{r, example3-expectations}
mu2d <- coenocline(locs, responseModel = "gaussian",
                   params = pars, extraParams = list(corr = 0.5),
                   expectation = TRUE)
```

`mu2d` now contains a matrix of expected species abundances, one column 
per species as before. Because of the way the `expand.grid()` function 
works, the ordering of species abudances in each column has the first 
gradient locations varying fastest --- the locations on the first 
gradient are repeated in order for each location on the second gradient

```{r, example3-head-locs}
head(locs)
```

As a result, we can reshape the abundances for a single species into a 
matrix reflecting the grid of locations over the gradient space via a 
simple `matrix()` call, setting the number of columns in the resultant 
matrix equal to the number of gradient locations in the simulation. By 
way of illustration, this approach is used to prepare the expected 
abundances for four of the species in `mu2d` for plotting via the 
`persp()` plotting function

```{r, example3-persp-plots, fig.cap = "Figure 5: Bivariate Gaussian species responses for four selected species."}
layout(matrix(1:4, ncol = 2))
op <- par(mar = rep(1, 4))
for (i in c(2,8,13,19)) {
    persp(loc1, loc2, matrix(mu2d[, i], ncol = length(loc2)),
          ticktype = "detailed", zlab = "Abundance",
          theta = 45, phi = 30)
}
par(op)
layout(1)
```

The selected species response curves are shown in Figure 5.

Simulated counts for each species can be produced by removing 
`expectation = TRUE` from the call and choosing an error distribution 
to make random draws from. For example, for negative binomial errors 
with dispersion \\(\alpha = 1\\) we can use

```{r, example3-nb-sim}
sim2d <- coenocline(locs, responseModel = "gaussian",
                    params = pars, extraParams = list(corr = 0.5),
                    countModel = "negbin", countParams = list(alpha = 1))
```

The resulting simulated counts for the same four selected species are 
shown in Figure 6, which was generated using the code below

```{r, example3-persp-plots2, fig.cap = "Figure 6: Simulated counts using negative binomial errors from bivariate Gaussian species responses for four selected species."}
layout(matrix(1:4, ncol = 2))
op <- par(mar = rep(1, 4))
for (i in c(2,8,13,19)) {
    persp(loc1, loc2, matrix(sim2d[, i], ncol = length(loc2)),
          ticktype = "detailed", zlab = "Abundance",
          theta = 45, phi = 30)
}
par(op)
layout(1)
```
