--- 
title: "User-friendly scaling"
status: publish
layout: post
published: true
type: post
tags:
- vegan
- biplot
- scaling
active: blog
category: R
---

```{r, echo = FALSE, include = FALSE, cache = FALSE}
render_jekyll()
opts_knit$set(progress = TRUE, base.url = "{{ site.url }}/assets/img/posts/", base.dir = "/home/gavin/work/web/jekyll/blog/assets/img/posts/")
opts_chunk$set(results = "hold",
               fig.path = "user-friendly-scaling-", 
               comment = NA,
               prompt = FALSE,
               cache = TRUE,
               cache.path = "../_knitr-cache/user-friendly-scaling-", 
               fig.show = "hold",
               dev = "png",
               dpi = 85,
               tidy = FALSE,
               fig.height = 7,
               fig.width = 7,
               dev.args = list(pointsize = 10))
opts_chunk$set(fig.lp = "fig:")
```

Back in the mists of time, whilst programming early versions of Canoco, 
Cajo ter Braak decided to allow users to specify how species and site 
ordination scores were scaled relative to one another via a simple 
numeric coding system. This was fine for the DOS-based software that 
Canoco was at the time; you entered `2` when prompted and you got 
*species* scaling, `-1` got you *site* or *sample* scaling **and** 
Hill's scaling or correlation-based scores depending on whether your 
ordination was a linear or unimodal method. This system persisted; even 
in the Windows era of Canoco these numeric codes can be found lurking 
in the `.con` files that describe the analysis performed. This use of 
numeric codes for scaling types was so pervasive that it was logical for 
Jari Oksanen to include the same system when the first `cca()` and 
`rda()` functions were written and in doing so Jari perpetuated one of 
the most frustrating things I've ever had to deal with as a user and 
teacher of ordination methods. But, as of last week, my frustration is 
no more...

...because we released a patch update to the CRAN version of **vegan**. 
Normally we don't introduce new functionality in patch releases but the 
change I made to the way users can request ordination scores was pretty 
trivial and maintained backwards compatibility.

Previously, different scalings could be requested using the `scaling` 
argument. `scaling` is an argument of the `scores()` function; anything 
function using `scores()` would either have `scaling` as a formal 
argument too, or would pass `scaling` on to `scores()` internally. To 
date, the different scores were specified as per DOS-era Canoco as 
numeric values. Now, `scores()` accepts either those same old numeric 
values or a character string for `scaling` coupled with a second 
logical argument. **Vegan** accepts the following character values to 
select the type of scaling:

 * `"sites"`, which gives site-focussed scaling, equivalent to numeric 
   value `1`
 
 * `"species"` (the default), which gives species- (variable-) focused 
   scaling, equivalent to numeric value `2`
 
 * `"symmetric"`, which gives a so-called symmetric scaling, and is 
   equivalent to numeric value `3`.

To get negative versions of these values, the `correlation` or `hill` 
argument should be set to `TRUE` as follows

 * `correlation` (default `FALSE`) for correlation-like scores for
   PCA/RDA/CAPSCALE models, or

 * `hill` (default `FALSE`) for Hill's scaling for CA/CCA models

Whilst this requires the setting of two different arguments, it's 
certainly a lot easier to remember these two arguments than what the 
numerical codes mean.

### Obligatory Dutch dune meadows example

Here's a quick example of the new usage showing a PCA of the classic 
Dutch dune meadow data set.

```{r example, fig = TRUE, fig.cap = "PCA of the Dutch dune meadow data set. Both biplots are drawn using species scaling, but the one on the right standardizes the species scores.", fig.width = 14}
library("vegan")
data(dune)

ord <- rda(dune)                  # fit the PCA
layout(matrix(1:2, ncol = 2))
plot(ord, scaling = "species")
plot(ord, scaling = "species", correlation = TRUE)
layout(1)
```

The two biplots are based on the same underlying ordination and both 
focus the scaling on best representing the relationships between 
species (`scaling = "species"`), but the biplot on the right uses 
correlation-like scores. This has the effect of making the species have 
equal representation on the plot without doing the PCA with 
standardized species data (all species having unit variance).
