--- 
title: "gratia 0.4.1 released"
date: 2020-05-31 13:00:00
subtitle: false
status: publish
layout: post
published: true
type: post
tags:
- GAM
- gratia
- R
- package
active: blog
category: R
twitterimg: "gratia-0-4-1-release-plot-difference-smooths-1.png"
---

```{r setup, echo = FALSE, include = FALSE, cache = FALSE}
## Change the `fig.path` and `cache.path` parts to reflect the new post
render_jekyll()
opts_knit$set(progress = TRUE, base.url = "{{ site.url }}/assets/img/posts/", base.dir = "/home/gavin/work/web/jekyll/blog/assets/img/posts/")
opts_chunk$set(results = "hold",
               fig.path = "gratia-0-4-1-release-", 
               comment = NA,
               prompt = FALSE,
               cache = FALSE,
               cache.path = "../_knitr-cache/gratia-0-4-1-release-", 
               fig.show = "hold",
               dev = "png",
               dpi = 85,
               tidy = FALSE,
               fig.height = 6,
               fig.width = 8,
               dev.args = list(pointsize = 10))
opts_chunk$set(fig.lp = "fig:")

## packages
library("gratia")
library("mgcv")
library("ggplot2")
library('dplyr')
library('tidyr')
```

After a slight snafu related to the 1.0.0 release of **dplyr**, a new version of **gratia** is out and available on CRAN. This release brings a number of new features, including differences of smooths, partial residuals on partial plots of univariate smooths, and a number of utility functions, while under the hood **gratia** works for a wider range of models that can be fitted by **mgcv**.

### Partial residuals

The `draw()` method for `gam()` and related models produces partial effects plots. `plot.gam()` has long had the ability to add partial residuals to partial plots of univariate smooths, and with the latest release `draw()` can now do so too.

```{r partial-residuals, fig.cap = "Partial plots of estimated smooth functions with partial residuals"}
df1 <- data_sim("eg1", n = 400, seed = 42)
m1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df1, method = "REML")
draw(m1, residuals = TRUE)
```

If the estimated functions have the correct degree of wiggliness, the partial residuals should be approximately uniformly distributed about the estimated smooth.

### Simulating data

The previous example demonstrated another new feature of the latest release; `data_sim()`. This is a reimplementation of `mgcv::gamSim()`, which is used to simulate data for testing GAMs. Data can be simulated from several widely-used functions that illustrate the power an capabilities of estimating smooth functions using penalised splines.

`data_sim()` returns simulated data in a tidy fashion and all the various example test data sets return consistently. Also, data from the example functions can be simulated from a number of probability distributions --- currently the Gaussian, Poisson, and Bernoulli distributions are supported, but future versions will offer a wider range to simulate from.

For example, the response data modelled above came from the following four functions used by Gu and Wahba

```{r data-sim, fig.cap = "Gu and Wahba four term additive example functions"}
df1 %>% mutate(id = seq_len(nrow(df1))) %>%
  select(id, x0:x3, f0:f3) %>%
  pivot_longer(x0:f3, names_sep = 1, names_to = c("var", "fun")) %>%
  pivot_wider(names_from = var, values_from = value) %>%
  ggplot(aes(x = x, y = f)) + 
    geom_line() + 
    facet_wrap(~ fun)
```

### Difference smooths

When GAMs contain smooth-factor interactions, we often want to compare smooths between levels of the factor to determine how the smooth effects vary between groups. The new release contains a function `difference_smooths()` that implements this idea.

The **mgcv** example for factor-smooth interactions using the `by` mechanism can be simulated from using `data_sim()`. The model fitted to the data contains a smooth of covariate `x1` and a smooth of `x2` for each level of the factor `fac`. Note that we need the parametric effect for `fac` as the `by` smooths are all centred about 0; the parametric term models the different group means.

```{r difference-smooths-model}
df <- data_sim("eg4", n = 1000, seed = 42)
m2 <- gam(y ~ fac + s(x2, by = fac) + s(x0), data = df, method = "REML")
```

`difference_smooths()` returns differences between the smooth functions for all pairs of the levels of `fac`, plus a credible interval for the difference.

```{r difference-smooths}
sm_diffs <- difference_smooths(m2, smooth = "s(x2)")
sm_diffs
```

There is a `draw()` method for objects returned by `difference_smooths()`, which will plot the pairwise differences

```{r plot-difference-smooths, fig.cap = "Differences between estimated smooth functions"}
draw(sm_diffs)
```

Note that these differences exclude differences in the group means and the differences between smooths are computed on the scale of the link function. A future version will allow for differences that include the group means.

### Fitted values and residuals utility functions

Two new utility functions are in the current release, `add_fitted()` and `add_residuals()` add fitted values and residuals to a data frame of observations used to fit a model.

```{r add-fitted-and-residuals}
df1 %>% add_fitted(m1, value = ".fitted") %>%
  add_residuals(m1, value = ".resid")
```

### Other changes

This release contains a number of other less-visible changes. **gratia** now handles models fitted by `gamm4::gamm4()` in more functions than before, while the utility functions `link()` and `inv_link()` now work for all families in **mgcv**, including the general family functions and those used for fitting location scale models.
