--- 
title: "gratia 0.9.0"
date: 2024-03-28 10:00:00
status: publish
layout: post
published: true
type: post
tags:
- gratia
- GAM
- "Generalized additive models"
- "R package"
- "R"
active: blog
category: R
twitterimg: "gratia-0-9-0-gratia-0-9-0-sos-example-1.png"
---

```{r setup, echo = FALSE, include = FALSE, cache = FALSE}
## Change the `fig.path` and `cache.path` parts to reflect the new post
render_jekyll()
opts_knit$set(progress = TRUE,
  base.url = "{{ site.url }}/assets/img/posts/",
  base.dir = "/home/au690221/work/web/jekyll/blog/assets/img/posts/")

opts_chunk$set(
  results = "hold",
  fig.path = "gratia-0-9-0-",
  comment = NA,
  prompt = FALSE,
  cache = TRUE,
  cache.path = "../_knitr-cache/gratia-0-9-0-",
  fig.show = "hold",
  dev = "png",
  dpi = 85,
  tidy = FALSE,
  fig.height = 6,
  fig.width = 8,
  dev.args = list(pointsize = 10))
opts_chunk$set(fig.lp = "fig:")
```

```{r package-setup-hidden, echo = FALSE, message = FALSE}
library('mgcv')
library('gratia')
library('dplyr')
```

I am pleased to announce the release of
[gratia](https://gavinsimpson.github.io/gratia/) 0.9.0. This release has been
over a year in the making and provides many new features as well as a more
consistent user experience. Unfortunately, I have had to make a lot of breaking
changes; nothing too egregious, but most user-facing functions are affected.
This release represents a solid base to move towards gratia version 1.0.0. Here,
I describe what I broke as well as outline some of the major new features in the
package.

## Breaking changes

Several of the main user-facing functions return user data along side the
variables created by those functions. For example, `smooth_estimates()` returns
the values of the covariates at which a smooth is evaluated.

```{r gratia-0-9-0-smooth-est-eg}
library("mgcv")
library("gratia")

df <- data_sim("eg1", n = 500, seed = 2)
m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = "REML")

smooth_estimates(m) # evaluate all smooths
```

`smooth_estimates()` returns info about the smooth, what is's label is, what
type it is, plus info relating to whether it is a factor by smooth. It isn't
unreasonable to think that a user might also use covariates with names `type`,
for example, which would immediately cause a problem if they fitted a model with
a `s(x, by = type)` smooth. We'd see a clash between the `type` that
`smooth_estimates()` wants to add to the return object and the variable the user
used which also needs to included in the return object.

Now, I *should* have foreseen this when I set about building gratia, but I
didn't, and so we arrive at March 2024 and the Great Renaming.

As you can see above, the variables that gratia's functions add to returned objects are now prefixed with a period: `.smooth` instead of `smooth`. I also took the opportunity to be more consistent and clear about what variables are through their naming. Hence, `smooth_estimates()` now returns a variable `.estimate` where previously I have `est`.

Some functions have changed more than others. `derivatives()`, for example, used to have a `data` variable that stored the covariate values at which the derivative of a smooth was computed. This wasn't very flexible unfortunately, and it wouldn't work `by` smooths as at some point we'd need to also store the by variable name and info and you can't easily stack factors with different levels without merging them. So now `derivatives()` more closely follows the conventions of `smooth_estimates()`

```{r gratia-0-9-0-derivatives-eg}
derivatives(m)
```

There are still some inconsistencies; no `.type` in the `derivatives()` output, but the `.fs` variable is present. Going forward, I'll be addressing these inconsistencies, but I'll be able to do them in a way that shouldn't break people's code.

I didn't make these changes lightly; I appreciate that these naming changes will cause code to break, not least a lot of my own. However, I truly believe that how things work now in 0.9.0 is the right way to combine user data with function-generated variables. Let's face it, if you name your variables with a `.` prefix, that's a you problem, not a me problem.

The other major change is in how spline-on-the-sphere (SOS) smooths are plotted. With version 3.5.0 of the ggplot2 package, the developers introduced a new guides system. I had been using `coord_map()` to generate a plot of an estimated SOS spline that looked like a sphere. Unfortunately, since I started using `coord_map()` the ggplot2 devs soft-deprecated the function and that meant that the new guide system wasn't applied to `coord_map()`, and the current gratia plot code was now generating warnings. So, I have switched to `coord_sf()`, which is much better all round, but the way projection information is supplied to the `coord` is very different. So gone are the `projection` and `orientation` in their place we have `crs`, `default_crs`, and `lims_method`.

```{r gratia-0-9-0-sos-example, fig.cap="Spline on the sphere example, showing estimated spatial effect on ocean surface chlorophyll a.", fig.height=4}
data(chl, package = "gamair")
m_chla <- bam(chl ~ s(lat, lon, bs = "sos"), data = chl, method = "fREML",
  discrete = TRUE)

draw(m_chla, crs = "+proj=wintri")
```

The current implementation isn't 100% finished; I need to be much more careful than I am in how I create the grid of points to evaluate the SOS spline at when it gets near to +/-90 degrees latitude or +/- 180 degrees longitude. I also need to figure out how to show as much of the smooth as is possible with a given projection. Notice in the lower left corner of the plot above how the high chlorophyll are is clipped a little.

## New features

This release of gratia brings a lot of new functionality. For the full details, see the (change log)[https://gavinsimpson.github.io/gratia/news/index.html#gratia-090]. Below I highlight some of the more important improvements.

`fitted_values()` has started to be able to handle location-scale-shape  families available in mgcv. I don't yet have complete coverage of all such families, but as of 0.9.0, supported families are `gaulss()`, `gammals()`, `gumbls()`, `gevlss()`, `shash()`, `twlss()`, and `ziplss()`. The `ocat()` family is also now supported.

Soap film smoothers created with `bs = "so"` are now supported with their own plotting method. Previously, gratia would draw the smooth as a standard bivariate smooth.

`response_derivatives()` is a new function to estimate derivatives on the response scale and compute uncertainties in the estimates using posterior sampling. This is enabled by new function `derivative_samples()`, which is what does the actual posterior sampling.

On a related note, all the posterior sampling function in gratia
    * `posterior_samples()`,
    * `fitted_samples()`,
    * `predicted_samples()`,
    * `derivative_samples()`,
    * `smooth_samples()`,
    * `simulate()`.
can now use the simple Metropolis Hastings sampler provided by mgcv, which instead of using a Gaussian approximation to the posterior, uses proposals from a Gaussian ot *t* distribution alternated with random walk proposals. And yes, `posterior_samples()` is a new function.

A new [vignette on posterior sampling](https://gavinsimpson.github.io/gratia/articles/posterior-simulation.html) adds to the package documentation. It describes how and what we are sampling in relation to GAMs, and includes an example of the benefits of using the Metropolis Hastings sampler in some situations.

`data_sim()` gains a bunch of new functionality and (slightly) better
documentation. The function can simulate data from a wider range of response
distributions than previously, and it also includes several new "models", known smooth effects, including data for use with mgcv's new `gfam()` family, which allows you to model responses of mixed type (continuous, binary, count, etc.)

`add_fitted_samples()`, `add_predicted_samples()`, `add_posterior_samples()`, and `add_smooth_samples()` are new utility functions that add the respective draws from the posterior distribution to an existing data object for the covariate values in that object: `obj |> add_posterior_draws(model)`.

```{r gratia-0-9-0-add-fitted-samples}
ds <- data_slice(m, x2 = evenly(x2, n = 20))
ds |>
  add_fitted_samples(model = m)
```

`draw.gam()` can now group factor by smooths for a given factor into a single panel, rather than plotting the smooths for each level in separate panels. This is achieved via new argument `grouped_by`.

```{r gratia-0-9-0-grouped-by, fig.cap="Plot produced by draw() using grouped_by = TRUE.", fig.height = 4}
df2 <- data_sim("eg4", seed = 2, n = 1000)
m2 <- gam(y ~ fac + s(x2, by = fac) + s(x0),
  data = df2, method = "REML")

m2 |>
  draw(grouped_by = TRUE)
```

For a full list of changes, see the (change log)[https://gavinsimpson.github.io/gratia/news/index.html#gratia-090].

## Defunct and deprecated

I have finally taken the decision to remove `evaluate_smooth()` from gratia. This function, alongside `fderiv()`, was the original functionality of the package from before it was even called gratia. It has long been superseded by `smooth_estimates()` however, and it became too difficult to maintain it.

This version of gratia also sees the deprecation of `evaluate_parametric_term()` and `datagen()`. The former is the counterpart to `evaluate_smooth()` but for parametric model terms; this has been superseded by `parametric_terms()`. `datagen()` was an early attempt at `data_slice()` and I never really used it as it wasn't very flexible or useful. These functions will be removed from gratia by version 0.11.0 or 1.0.0, whichever of those happens first.

## Fin

Version 0.9.0 of gratia is now on CRAN. I hope you find the new version of gratia useful and can bear the annoyances of code breaking. If you have thoughts about the new release, what could be improved and changed, let me know in the comments or in a GitHub Issue.

